The emotion--facial expression link: Evidence from human and automatic expression recognition

### 2020-11-04 manuscript/manuscript_apa.tex
acknowledgment

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  doc]{apa7}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The emotion--facial expression link: Evidence from human and automatic expression recognition},
  pdfauthor={Anna Tcherkassof1 \& Damien Dupré2},
  pdflang={en-EN},
  pdfkeywords={emotion, facial expression, self-report, human observer, automatic recognition.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{The emotion facial expression link}
\keywords{emotion, facial expression, self-report, human observer, automatic recognition.\newline\indent Word count: 7943}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}
\usepackage{wrapfig}
\usepackage[none]{hyphenat}
\floatplacement{figure}{H}
\usepackage{flushend}
\usepackage{biblatex}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

@@

@@

\abstract{
While it has been taken for granted in the development of several automatic facial expression recognition tools, the question of the coherence between subjective feelings and facial expressions is still a subject of debate. On one hand, the ``Basic Emotion View'' conceives emotions as genetically hardwired and therefore being genuinely displayed through facial expressions. On the other hand, the constructivist approach conceives emotions as socially constructed; the emotional meaning of a facial expression being inferred by the observer. In order to evaluate the coherence between the subjective feeling of emotions and their recognition based on facial expression, 232 videos of encoders recruited to carry out an emotion elicitation task were annotated by 1383 human observers as well as by an automatic facial expression classifier. Results show low accuracy of human observers and of the automatic classifier to infer the subjective feeling from the facial expressions displayed by encoders. They also show a weak consistency between self-reported emotional states and facial emotional displays. Based on these results, the hypothesis of genetically hardwired emotion genuinely displayed is difficult to support, whereas the idea of emotion and facial expression as being socially constructed appears to be more likely. Accordingly, automatic emotion recognition tools based on facial expressions should be questioned.
}

@@

With the development of commercial automatic facial expression recognition tools (Dupré et al., 2020, 2018), industries and governments are gradually implementing this technology in order to monitor humans' emotions in various scenarios (\emph{e.g.}, in marketing, healthcare, and the automotive industry to name a few). This technology rests on the premise that facial expressions provide direct access to individuals' subjective feelings, and that one can read the emotions displayed on a person's face as one would an open book. This premise is derived from the ``Basic Emotion View'', a theory which suggests a one-to-one mapping between subjective feeling and facial expression (Ekman et al., 1987; Ekman, 1992, 2007; Ekman \& Heider, 1988). Even if this premise is central to the modern mainstream approach of human emotion, recent research in affective science is challenging it. Once the Basic Emotion View is briefly described, its foremost criticisms will be exposed. Inconsistencies and unpredicted findings come both from field observations and laboratory experiments. They relate to two different issues. The first issue is the sender's production of spontaneous facial displays. The second issue relates to the receiver's interpretation of these facial displays. Building on this evidence, and as an attempt to shed further light on the emotion--face link, the present study's main goal is to tackle these two issues at the same time. It aims at identifying the extent to which ordinary people's experienced emotions are displayed through identifiable Emotional Facial Expressions (EFE) recognized as such both by human individuals, and by automatic facial expression recognition tools.

@@

The belief that facial expression is linked to emotional states can be traced back to Darwin in \emph{The Expression of the Emotions in Man and Animals} (Darwin, 1872). Countless studies have fortified the idea that emotional states are inherently coupled to a set of prototypic facial expressions (\emph{e.g.}, Ekman \& Cordaro, 2011). The Basic Emotion View holds that facial expressions are genuine displays of an individual's inner emotional state. More specifically, a set of six emotions (\emph{fear}, \emph{anger}, \emph{surprise}, \emph{disgust}, \emph{sadness} and \emph{joy}) are universally displayed and are genetically hardwired not only in humans (Ekman, 1992), but also in different animal species (Waal, 2019). According to the Basic Emotion View, \emph{``when emotions are aroused by perception of a social event, a set of central commands produce patterned emotion-specific changes in multiple systems, including {[}\ldots{]} facial expressions.''} (Ekman, 2007, p. 49). To respond to criticisms, several amendments have been made to the Basic Emotion View, increasing the number of basic emotions from six to seven (Ekman \& Heider, 1988) as well as adding the concept of ``display rules'' to explain cultural differences in the management of facial expressions (Ekman et al., 1987).

Relatively few studies have actually tested Basic Emotion View's fundamental claim regarding the facial production (so-called ``expression'') of emotions. Rosenberg \& Ekman (1994) reported the first evidence of coherence between self-report of emotion and displayed facial expressions. Participants were shown emotionally evocative films and reported their own emotions using a moment-by-moment reporting procedure. Analysis of participants' facial expressions and reports of emotions showed that there was a high degree of temporal linkage and emotional agreement between facial expressions and self-reports. Notwithstanding this study, few have used purely descriptive methods such as electromyography or objective face coding systems to identify and to measure the actual changes in the face when a given emotion is felt (see Wagner, 1997, for methodological points). Much of the research has focused instead on the recognition of EFE, that is on the issue of the viewer's interpretation of facial displays. As people have been thought to display their internal state, EFE supposedly serve as a window into the emotions of others. Viewers ``recognize'' the facial displays and infer the corresponding emotional state. The Readout Hypothesis (Buck, 1985) formalizes this folk theory. According to it, facial expressions have evolved to provide \emph{``an external readout of those motivational-emotional processes that have had social implications during the course of evolution''} (Buck, 1985, pp. 396--397). Thus, as readouts of emotional states, spontaneous expressive displays are directly accessible to other organisms. Research on the inference of emotions from facial expressions has established that viewers show considerable agreement that the so-called basic emotions---happiness, anger, fear, disgust, sadness, and surprise---are associated with specific facial displays (\emph{e.g.}, reviews from Elfenbein \& Ambady, 2002; Russell, 1994).

Yet, in spite of the popular support it has received, the empirical data called upon by the Basic Emotion View remains unpersuasive. Notably, it fails to explain, in instances in which display rules cannot be called upon, how individuals can feel emotions without expressing them or how individuals can express emotions without feeling them, and why observers can make mistakes in recognizing facial expressions of basic emotions, among others. The following is a summary of key findings and conclusions both from field observations and laboratory experiments on spontaneous expression of emotions, all of them strongly subverting the Basic Emotion viewpoint (for reviews, Fernández-Dols \& Crivelli, 2013; Reisenzein et al., 2013; Russell et al., 2003). Mains issues regarding recognition studies are also briefly discussed afterwards.

@@

Naturalistic studies look at the ecological frequency of co-occurrence of certain emotions and facial displays (Fernández-Dols \& Crivelli, 2013). Though they cannot be considered as a straightforward test of the triggering role of emotion on facial behavior, they have the advantage of considering situations that cannot be created in a laboratory. For instance, the ethological study of Kraut \& Johnston (1979) contradicts the Basic Emotion View premise that a smile is only the major component of a facial display associated with and caused by feelings of joy or happiness. Naturalistic observation at a bowling alley showed that bowlers do not necessarily smile after scoring a spare or a strike (a situation likely to elicit a positive emotion). Rather, they often smile when interacting with other people. More generally, examination of bowlers' facial display showed that they rarely smiled while facing the pins but often smiled when facing their friends. These findings were confirmed by Ruiz-Belda et al. (2003) who analyzed the facial displays of bowlers after a strike and soccer fans after their team scored. Their results show a low probability of smiling when participants were not interacting with someone else. These findings have been supported in other realistic field settings such as the one of Fernández-Dols \& Ruiz-Belda (1995) suggesting that happiness is not a sufficient cause of smiling. Fernández-Dols \& Ruiz-Belda (1995) watched extremely happy gold medalist athletes displaying facial expressions of sadness (sometimes associated with tears) during their Olympics awards ceremonies. More specifically, winners showed Duchenne smiles and other types of smiles when they were interacting with other people during the awards ceremony, but smiles were scarce or nonexistent when waiting behind the podium and/or when turning toward the flagpoles and focusing their attention on the flags and the national anthem. Crivelli et al. (2015) also found that the strongest predictor for the occurrence of Duchenne smiles in judo winners is when they are engaged in social interaction and not when they just won their match.

Happiness/joy is not the only emotion weakly associated with the predicted facial expression. A naturalistic study conducted by Scherer \& Ceschi (1997) in an airport's baggage handling office showed that passengers claiming for their lost luggage displayed very few facial expressions of negative emotions while self-reporting subjective feeling states of anger or sadness among others. The covariation between passengers' self-ratings and the claims agents' attributions of the passengers' emotions was very low. Another refutation of the Basic Emotion View's predictions regarding the link between emotions and facial expressions is provided by naturalistic observations of infants' productions of facial expressions. It is often acknowledged that adults regulate their expressive behavior. In accordance with various display rules (personal, social, cultural ones), they exert a control over the supposed automatic readouts of their emotions. Developmental studies provide relevant settings in which display rules are inoperative. Camras and her colleagues observed that facial expressions of negative emotions were displayed in circumstances that were unlikely to have elicited those emotions. For instance, infants displayed ``fear'' expressions in settings not related to fear (Camras et al., 1991). Bennett et al. (2002) videotaped a sample of 4-month-old infants during tickle, sour taste, jack-in-the-box, arm restraint, and masked stranger situations. Infants displayed a variety of facial expressions in each eliciting situation. Yet, more infants exhibited positive than negative facial expressions across all situations --except sour taste. No evidence for emotion-specific facial expressions corresponding to anger, fear, and sadness was obtained. Camras and her colleagues also observed that 11-month-old European-American, Chinese, and Japanese infants did not display distinct negative emotion-specific patterns of facial muscles in response to two elicitors meant to induce fear and anger (Camras et al., 2007). Thus, these findings in naturalistic settings provide little support for the one-to-one mapping of subjective feeling and facial expression.

@@

The Basic Emotion View postulates that, when triggered, each basic emotion is expressed by a prototypical face (non-basic emotions being blends of the basic ones). In return, the recognition of EFE is claimed to be based on the identification of specific patterns of facial movements associated with each emotion, as if expression and recognition were the two sides of the same coin. Seemingly compelling evidence supports this claim, sustaining the possibility of a clear readout of subjective feelings from facial expressions. Furthermore, recognition systems rely on such a principle and are considered as an objective coding tool because they are based on the identification of specific muscular changes in the face. Many people strongly agree that so-called basic emotions are associated with specific facial configurations (Ekman, 2017; Elfenbein \& Ambady, 2002) and also argue that this is strong evidence for the Basic Emotion View. Moreover, it implies that EFE recognition of human observers should be as accurate as automatic classifiers'.

@@

To the methodological limitations contaminating the hundreds of studies apparently supporting the Basic Emotion View, a stimulus bias must be added. Facial stimuli used in experiments also constitute a methodological bias because they are unrepresentative of ordinary facial expressions. Basic Emotion View empirical evidence is based for the most part on methods using a static and unnatural material, namely, still photographs of posed facial expressions of emotion (\emph{e.g.}, intentionally encoded by the sender). This kind of methodology raises questions about its ecological validity and the generalizability of the results to real interpersonal emotional communication (\emph{e.g.}, Tcherkassof et al., 2007). Indeed, a number of pieces of evidence indicate that research cannot content itself with data collected with static and posed material. These data come from researches studying the case of dynamic and/or spontaneous facial expressions of emotion. They show that the dynamic aspects of facial movement are likely to be of importance (\emph{e.g.}, Kamachi et al., 2013). Cohn \& Schmidt (2004) have shown that spontaneous smiles are of smaller amplitude and have a more consistent relation between amplitude and duration than deliberate smiles. Hess \& Kleck (1990) have also pointed out the importance of the dynamics of facial movements, and particularly the irregularity, or phasic changes, of the expressions' unfolding. Thus, the motion of facial expression provides observers with other information than the one provided by static expressions. It may be that differences in the social information displayed by static and dynamic expressions leads to facial recognition differential effects. Regarding the issue of spontaneous \emph{vs.} posed expressions (the latter are overused in experiments), as Meillon et al. (2010) conclude, EFE have been typically studied as static displays. As a consequence, even though the central role of the dynamics of facial expressions is endorsed, little is still known about the temporal course of facial expressions. Furthermore, studied EFE exhibit emotions simulated or posed by actors. Yet, the lack of spontaneity and naturalness of this material constitutes a serious objection raised against such studies (Kanade et al., 2000).

@@

To evaluate the consistency between subjective feeling of emotions and their recognition from facial expressions, encoders were first recruited to perform an emotion elicitation task while their facial expression was video recorded. In order to reduce the likelihood of facial control, the encoders were alone in the room and were filmed by hidden cameras, so they had no reason to comply with social display rules. Then, the videos of the encoders' faces were shown to human observers and were also analyzed by an automatic classifier in order to identify which emotion was displayed.

@@

\hypertarget{human-observers-accuracy}{%
\subsection{Human Observers' Accuracy}\label{human-observers-accuracy}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa7}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The emotion--facial expression link: Evidence from human and automatic expression recognition},
  pdfauthor={Anna Tcherkassof1 \& Damien Dupré2},
  pdflang={en-EN},
  pdfkeywords={emotion, facial expression, self-report, human perceiver, automatic recognition.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{The emotion facial expression link}
\keywords{emotion, facial expression, self-report, human perceiver, automatic recognition.\newline\indent Word count: 8558}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}
\usepackage{wrapfig}
\usepackage[none]{hyphenat}
\floatplacement{figure}{H}
\usepackage{flushend}
\usepackage{biblatex}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

@@

\authornote{

We thank anonymous reviewers for their insightful comments on a previous version of the manuscript.

The authors made the following contributions. Anna Tcherkassof: Conceptualization, Project administration, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Damien Dupré: Investigation, Data curation, Formal Analysis, Software, Visualization, Writing - Original Draft Preparation, Writing - Review \& Editing.

Correspondence concerning this article should be addressed to Anna Tcherkassof, Bâtiment Michel Dubois, Université Grenoble Alpes, 1251 Avenue Centrale, 38400 Saint-Martin-d'Hères. E-mail: \href{mailto:anna.tcherkassof@univ-grenoble-alpes.fr}{\nolinkurl{anna.tcherkassof@univ-grenoble-alpes.fr}}

}

@@

\abstract{
While it has been taken for granted in the development of several automatic facial expression recognition tools, the question of the coherence between subjective feelings and facial expressions is still a subject of debate. On one hand, the ``Basic Emotion View'' conceives emotions as genetically hardwired and therefore being genuinely displayed through facial expressions. Consequently, emotion recognition is perceiver-independent. On the other hand, the constructivist approach conceives emotions as socially constructed, the emotional meaning of a facial expression being inferred by the perceiver. Hence, emotion recognition is perceiver-dependent. In order 1) to evaluate the coherence between the subjective feeling of emotions and their spontaneous facial displays, and 2) to compare the recognition of such displays by human perceivers and by an automatic facial expression classifier, 232 videos of expressers recruited to carry out an emotion elicitation task were annotated by 1383 human perceivers as well as by Affdex, an automatic classifier. Results show a weak consistency between self-reported emotional states by expressers and their facial emotional displays. They also show low accuracy both of human perceivers and of the automatic classifier to infer the subjective feeling from the spontaneous facial expressions displayed by expressers. However, the results are more in favor of a perceiver-dependent view. Based on these results, the hypothesis of genetically hardwired emotion genuinely displayed is difficult to support, whereas the idea of emotion and facial expression as being socially constructed appears to be more likely. Accordingly, automatic emotion recognition tools based on facial expressions should be questioned.
}

@@

With the development of commercial automatic facial expression recognition tools (Dupré et al., 2020, 2018), industries and governments are gradually implementing this technology in order to monitor humans' emotions in various scenarios (\emph{e.g.}, in marketing, healthcare, and the automotive industry to name a few). This technology rests on the premise that emotional state is revealed by the measurement of facial features thanks to a one-to-one mapping between subjective feeling and facial expression. This premise comes from the Basic Emotion View (Ekman et al., 1987; Ekman, 1992, 2007; Ekman \& Heider, 1988). By contrast, for the constructivist approach (barrett2017emotions; Gendron \& Barrett, 2017), emotions being socially constructed, a making-meaning process takes part in the link between emotion and facial expression. Emotion recognition is considered as resulting from the perceiver making inferences about the expresser's underlying emotions. The constructivist approach is therefore a perceiver-\emph{dependent} view whereas the Basic Emotion View is a perceiver-\emph{independent} view (Gendron \& Barrett, 2017). In the emotion--facial expression link contention, the issue of facial expressions' naturalness must also be considered. Since automatic recognition tools are meant to be used in real-life conditions, spontaneous displays of emotional feelings are the heart of the matter as well as their accurate recognition. The potential significance of the present study is that it aims at providing data, not only on the emotion--facial expression link, but also on the perceiver-independent \emph{vs.} perceiver-dependent question of emotion recognition as regard to spontaneous facial expressions of self-reported emotions. It looks at the triangulation emotional feeling---spontaneous facial display---emotion recognition within a single paradigm comparing the recognition accuracy between human and machines.

@@

Since Darwin's book \emph{The Expression of the Emotions in Man and Animals} (Darwin, 1872), countless studies have fortified the idea that emotional states are inherently coupled to a set of prototypic facial expressions (\emph{e.g.}, Ekman \& Cordaro, 2011). The Basic Emotion View holds that facial expressions are genuine displays of an individual's inner emotional state. More specifically, a set of six so-called basic emotions (\emph{fear}, \emph{anger}, \emph{surprise}, \emph{disgust}, \emph{sadness} and \emph{happiness}) are universally displayed and are genetically hardwired not only in humans (Ekman, 1992), but also in different animal species (Waal, 2019). According to the Basic Emotion View, \emph{``when emotions are aroused by perception of a social event, a set of central commands produce patterned emotion-specific changes in multiple systems, including {[}\ldots{]} facial expressions.''} (Ekman, 2007, p. 49). To respond to criticisms, several amendments have been made to the Basic Emotion View, increasing the number of basic emotions from six to seven (Ekman \& Heider, 1988) as well as adding the concept of ``display rules'' to explain cultural differences in the management of facial expressions (Ekman et al., 1987). The most recent versions of the Basic Emotion Theory even distinguish some twenty expressions of emotions, including love, pride, embarrassment, shame, among others (Keltner et al., 2019).

Relatively few studies have actually tested Basic Emotion View's fundamental claim regarding the facial production (or ``expression'') of emotions. Rosenberg \& Ekman (1994) reported the first evidence of coherence between self-report of emotion and displayed facial expressions. Participants were shown emotionally evocative films and reported their own emotions using a moment-by-moment reporting procedure. Analysis of participants' facial expressions and reports of emotions showed that there was a high degree of temporal linkage and emotional agreement between facial expressions and self-reports. Notwithstanding this study, few have used purely descriptive methods such as electromyography or objective face coding systems to identify and to measure the actual changes in the face when a given emotion is felt (see Wagner, 1997, for methodological points).

Much of the research has focused instead on the recognition of emotional facial expressions (EFE), that is on the issue of the perceiver's interpretation of facial displays. According to the Readout Hypothesis (Buck, 1985) which formalizes the basic Emotion View, facial expressions have evolved to provide \emph{``an external readout of those motivational-emotional processes that have had social implications during the course of evolution''} (Buck, 1985, pp. 396--397). Thus, as readouts of emotional states, spontaneous expressive displays are directly accessible to other organisms. Research on the inference of emotions from facial expressions has established that perceivers show considerable agreement that the so-called basic emotions are associated with specific facial displays (\emph{e.g.}, reviews from Elfenbein \& Ambady, 2002; Russell, 1994).

Yet, the empirical data called upon by the Basic Emotion View remains unpersuasive. Notably, it fails to explain, in instances in which display rules cannot be called upon, how individuals can feel emotions without expressing them or how individuals can express emotions without feeling them, and why perceivers can make mistakes in recognizing facial expressions of basic emotions, among others. All these findings subverting the Basic Emotion viewpoint come from field observations and laboratory experiments on spontaneous expression of emotions (for reviews, Fernández-Dols \& Crivelli, 2013; Reisenzein et al., 2013; Russell et al., 2003). The key findings are summarized below as well as the main issues regarding recognition studies.

@@

Naturalistic studies look at the ecological frequency of co-occurrence of certain emotions and facial displays (Fernández-Dols \& Crivelli, 2013). Though they cannot be considered as a straightforward test of the triggering role of emotion on facial behavior, they have the advantage of considering situations that cannot be created in a laboratory. For instance, the ethological study of Kraut and Kraut \& Johnston (1979) contradicts the Basic Emotion View premise that a smile is only the major component of a facial display associated with and caused by feelings of happiness. Naturalistic observation at a bowling alley showed that bowlers do not necessarily smile after scoring a spare or a strike (a situation likely to elicit a positive emotion). Rather, they often smile when interacting with other people. More generally, examination of bowlers' facial display showed that they rarely smiled while facing the pins but often smiled when facing their friends. These findings were confirmed by Ruiz-Belda et al. (2003) who analyzed the facial displays of bowlers after a strike and soccer fans after their team scored. Their results show a low probability of smiling when participants were not interacting with someone else. These findings have been supported in other realistic field settings such as the one of Fernández-Dols \& Ruiz-Belda (1995) suggesting that happiness is not a sufficient cause of smiling. Fernández-Dols \& Ruiz-Belda (1995) watched extremely happy gold medalist athletes displaying facial expressions of sadness (sometimes associated with tears) during their Olympics awards ceremonies. More specifically, winners showed Duchenne smiles and other types of smiles when they were interacting with other people during the awards ceremony, but smiles were scarce or nonexistent when waiting behind the podium and/or when turning toward the flagpoles and focusing their attention on the flags and the national anthem. Other naturalistic studies have also found that happiness cannot be considered the strongest predictor for the occurrence of Duchenne smiles (Crivelli et al., 2015).

Happiness is not the only emotion weakly associated with the predicted facial expression. A naturalistic study conducted by Scherer \& Ceschi (1997) in an airport's baggage handling office showed that passengers claiming for their lost luggage displayed very few facial expressions of negative emotions while self-reporting subjective feeling states of anger or sadness among others. The covariation between passengers' self-ratings and the claims agents' attributions of the passengers' emotions was very low. Another refutation of the Basic Emotion View's predictions regarding the link between emotions and facial expressions is provided by naturalistic observations of infants' productions of facial expressions. It is often acknowledged that adults regulate their expressive behavior. In accordance with various display rules (personal, social, cultural ones), they exert a control over the supposed automatic readouts of their emotions. Developmental studies provide relevant settings in which display rules are inoperative. Camras and her colleagues observed that facial expressions of negative emotions were displayed in circumstances that were unlikely to have elicited those emotions. For instance, infants displayed ``fear'' expressions in settings not related to fear (Camras et al., 1991). Bennett et al. (2002) videotaped a sample of 4-month-old infants during tickle, sour taste, jack-in-the-box, arm restraint, and masked stranger situations. Infants displayed a variety of facial expressions in each eliciting situation. Yet, more infants exhibited positive than negative facial expressions across all situations --except sour taste. No evidence for emotion-specific facial expressions corresponding to anger, fear, and sadness was obtained. Camras and her colleagues also observed that 11-month-old European-American, Chinese, and Japanese infants did not display distinct negative emotion-specific patterns of facial muscles in response to two elicitors meant to induce fear and anger (Camras et al., 2007). Thus, these findings in naturalistic settings provide little support for the one-to-one mapping of subjective feeling and facial expression.

@@

The Basic Emotion View postulates that, when triggered, each basic emotion is expressed by a prototypical face (non-basic emotions being blends of the basic ones). In return, the recognition of EFE is claimed to be based on the identification of specific patterns of facial movements associated with each emotion, as if expression and recognition were the two sides of the same coin. Seemingly compelling evidence supports this claim, sustaining the possibility of a clear readout of subjective feelings from facial expressions. Furthermore, recognition systems rely on such a principle and are considered as an objective coding tool because they are based on the identification of specific muscular changes in the face. Many people strongly agree that so-called basic emotions are associated with specific facial configurations (Ekman, 2017; Elfenbein \& Ambady, 2002) and also argue that this is strong evidence for the Basic Emotion View. Moreover, it implies that EFE recognition of human perceivers should be as accurate as automatic classifiers.

@@

To the methodological limitations contaminating the hundreds of studies apparently supporting the Basic Emotion View, a stimulus bias must be added. Facial stimuli used in experiments also constitute a methodological bias because they are unrepresentative of ordinary facial expressions. Basic Emotion View empirical evidence is based for the most part on methods using a static and unnatural material, namely, still photographs of posed facial expressions of emotion (\emph{e.g.}, intentionally encoded by the expresser). This kind of methodology raises questions about its ecological validity and the generalizability of the results to real interpersonal emotional communication (\emph{e.g.}, Tcherkassof et al., 2007). Indeed, a number of pieces of evidence indicate that research cannot content itself with data collected with static and posed material. These data come from researches studying the case of dynamic and/or spontaneous facial expressions of emotion. They show that the dynamic aspects of facial movement are likely to be of importance (\emph{e.g.}, Kamachi et al., 2013). Cohn \& Schmidt (2004) have shown that spontaneous smiles are of smaller amplitude and have a more consistent relation between amplitude and duration than deliberate smiles. Hess \& Kleck (1990) have also pointed out the importance of the dynamics of facial movements, and particularly the irregularity, or phasic changes, of the expressions' unfolding. Thus, the motion of facial expression provides perceivers with other information than the one provided by static expressions. It may be that differences in the social information displayed by static and dynamic expressions leads to facial recognition differential effects. Regarding the issue of spontaneous \emph{vs.} posed expressions (the latter are overused in experiments), as Meillon et al. (2010) conclude, EFE have been typically studied as static displays. As a consequence, even though the central role of the dynamics of facial expressions is endorsed, little is still known about the temporal course of facial expressions. Furthermore, studied EFE exhibit emotions simulated or posed by actors. Yet, the lack of spontaneity and naturalness of this material constitutes a serious objection raised against such studies (Kanade et al., 2000).

@@

To evaluate the consistency between subjective feeling of emotions and their recognition from facial expressions, expressers were first recruited to perform an emotion elicitation task while their facial expression was video recorded. In order to reduce the likelihood of facial control, the expressers were alone in the room and were filmed by hidden cameras, so they had no reason to comply with social display rules. Then, the videos of the expressers' faces were shown to human perceivers and were also analyzed by an automatic classifier in order to identify which emotion was displayed.

@@

\hypertarget{human-perceivers-accuracy}{%
\subsection{Human Perceivers' Accuracy}\label{human-perceivers-accuracy}}
